### 2010-09-29 doc/guide/language.tex
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@3418 3e02f596-3131-0410-8352-e04eeaec37cc

\index{Statements!Rules}
\index{Statements!Facts}
\index{Statements!Integrity Constraints}
\begin{tabular}{rl@{}l}\label{eq:normal:rule}
\textbf{Rule:} & $A_0$& $\code{:-}~L_1\code{,}\dots\code{,}L_n$\code{.}
\\
\textbf{Fact:} & $A_0$&\code{.}
\\
\textbf{Integrity Constraint:} & & $\code{:-}~L_1\code{,}\dots\code{,}L_n$\code{.}
\end{tabular}
\\

\noindent
The head~$A_0$ of a fact or a rule is an \emph{atom} of the same form
as a function symbol or constant.
Any $L_j$ is a \emph{literal} of the form $A$ or $\code{not}~A$
for an atom~$A$ where the connective \code{not} corresponds to default negation\index{Literals!Default Negation}.
The set of literals $\{L_1,\dots,L_n\}$ is called the body of the rule.
Facts have an empty body.
Throughout this section we further extend the predicates that can be used 
in a rule including comparison predicates (Section \ref{subsec:gringo:comp}) and aggregates (Section \ref{subsec:gringo:aggregate}).
Furthermore, \gringo{} expects rules to be safe, i.e., 
all variables that appear in a rule have to appear in some positive literal (a literal not preceded by \code{not}) in the body.
If a variable appears positively in some predicate, then we say that this predicate binds the variable.

Intuitively, the head of a rule has to be true whenever all its body literals are true.
In ASP every atom needs some derivation, i.e., 
an atom cannot be true if there is no rule deriving it.
This implies that only atoms appearing in some head can appear in answer sets.
Furthermore, derivations%
\footnote{There are extensions like disjunctions 
that go beyond simple derivability and also require minimality w.r.t. a reduct. 
We do not cover the semantics of such constraints in this guide.}%
have to be acyclic,
a feature that is important to model reachability.
As a simple example, consider the program \code{a :- b. b :- a.}
The only answer set to this program is the empty set.
Adding either \code{a.} or \code{b.} to the program results in the answer set $\{a,b\}$.
Finally, note that default negation is ignored when checking for acyclic derivations 
(we do not need a reason for an atom being false).
Default negation can be used to express choices, e.g., the program \code{a :- not b. b :- not a.} 
has the two answer sets $\{a\}$ and $\{b\}$.
But in practice it is never needed to express choices this way.
For example in the introductory example in Section~\ref{sec:quickstart} we used a cardinality 
constraint, which provides a much more readable way to introduce choices.
\com{This paragraph contains a lot of information. Could someone help me making it more readable?}

@@

In logic programs, connective \code{not} expresses default negation,
that is, a literal $\code{not}~A$ is assumed to hold unless~$A$ is derived.
In contrast, the classical (or strong) negation of some proposition
holds if the complement of the proposition is objectively derived~\cite{gellif91a}.
Classical negation, indicated by symbol ``\code{-},'' is permitted in front of atoms.
That is, if~$A$ is an atom, then~$\code{-}A$ is the complement of~$A$.
Semantically, $\code{-}A$ is simply a new atom,
with the additional condition that~$A$ and~$\code{-}A$ must not jointly hold.
Observe that classical negation is merely a syntactic feature that can be
implemented via integrity constraints whose effect is to eliminate
any answer set candidate containing complementary atoms.

@@

It is important to note that variables
in the scope of an arithmetic function are not bound by a corresponding atom.
For instance, the rule \code{p(X) :- p(X+1).} is not safe but \code{p(X-1) :- p(X).} is.
Although, the latter might produce an infinite grounding 
and \gringo{} not necessarily halts when given such an input.

@@

All comparison predicates can also be used with arbitrary ground terms,
as in the next program:%
\marginlabel{%
  As above, invoking:\\
  \code{\mbox{~}gringo -t \attach{examples/symbc.lp}{symbc.lp}}
  yields the unique answer set of the program in terms of facts.}
%
\lstinputlisting{examples/symbc.lp}
%
Integers are compared in the usual way and constants are ordered lexicographically.
Function symbols are compared first using their arity.
If the arity differs, then the name of the function symbol is compared lexicographically.
If again the name differs, then arguments are compared component wise.
Finally, integers are always smaller than constants and constants are always smaller than function symbols.
\eexample
\comment{add \#sup and \#inf?}
\end{example}

Importantly, a built-in comparison predicate cannot bind variables, i.e.,
when checking whether a rule is safe, 
comparison predicates are not considered to be positive.

@@

Finally, note that assignments to some extent can bind variables.
Of course cyclic assignments cannot bind variables.
For example the rule \code{p(X) :- X = Y, Y = X.} is rejected by \gringo{}.
Either \code{X} or \code{Y} has to be provided by some positive predicate in this case.
Additionally, unification is restricted to ground terms on the right hand side of the assignment, 
that is, all variables on the right hand side have to be bound by some other predicate.

@@

As with built-in arithmetic functions,
an integer interval mentioning some variable (like~\var{X} in Line~4 of Example~\ref{ex:int})
cannot be used to bind the variable.

@@

Note that there are three important issues about the correct usage of conditions:
\begin{enumerate}
\item All predicates of atoms on the right-hand side of a condition
must be either domain predicates\comment{refer to appendix?},
i.e., predicates that can be completely evaluated during grounding,
or built-in, which is due to the fact that conditions are evaluated during grounding.
\item Any variable occurring within a condition is considered as \emph{local},
that is, a condition cannot be used to bind variables outside the condition.
In turn, variables outside conditions are \emph{global}, and each variable
within an atom in front of a condition must occur on the right-hand side or
be global.
\item Global variables take priority over local ones, that is,
they are instantiated first.
As a consequence, a local variable that also occurs globally is substituted by a term
before the ground instances of a condition are determined.
Hence, the names of local variables must be chosen with care,
making sure that they do not accidentally match the names of global variables.
\end{enumerate}

@@

\com{Should we mention stratifed aggregates that do not necessarily need conditions?}

@@

\paragraph{Constant Replacement.}
\index{Meta-Statements!Constant Replacement, \code{\#const}}%
Constants appearing in a logic program may actually be placeholders for
concrete values to be provided by a user.
An example of this is given in Section~\ref{subsec:ex:color}.
Via the \code{\#const} declarative,
one may define a default value to be inserted for a constant.
Such a default value can still be overridden via command line option
\code{--const} (cf.\ Section~\ref{subsec:opt:gringo}).
Syntactically, \code{\#const} must be followed by an assignment having
a (symbolic) constant on the left-hand side and a term on the right-hand side.
Some exemplary constant declarations are:
%
\begin{lstlisting}[numbers=none]
#const x = 42.
#const y = f(x,h).
\end{lstlisting}%
Note that (for efficiency reasons) constant declarations are order dependent.
In the example above \code{x} would be replaced by \code{42}
but when reversing the directives this would no longer be the case.

\paragraph{Domain Declarations.}
\index{Meta-Statements!Domain Declarations, \code{\#domain}}%
Usually, variable names are local to a rule,
where they must be bound via appropriate atoms.
This locality can be undermined by using
\code{\#domain} declarations 
that globally associate variable names to atoms.
An associated atom is then simply added to the body of a rule in which
such a predefined variable name occurs in.
The following is a made-up example:
%
\begin{lstlisting}[numbers=none]
p(1,1). p(1,2).
#domain p(X,Y).
#domain p(Y,Z).
q(Z,X) :- not p(Z,X).
\end{lstlisting}
%
The above program is a priori not safe because
variables~\var{X} and~\var{Z} are unbound in the last rule.
However, as they belong to \code{\#domain} declarations,
\gringo\ and \clingo\ expand the last rule to:
%
\begin{lstlisting}[numbers=none]
q(Z,X) :- p(X,Y), p(Y,Z), not p(Z,X).
\end{lstlisting}
%
Observe that the resulting program is safe.
Nevertheless, we suggest not to use domain statements because 
in ASP it is common to use very short variable names and
using domain statements likely results in name clashes
and undesired behavior.

@@

@@

\marginlabel{%
Try to invoke this program multiple times using:\\
\code{%
\mbox{~}clingo \attach{examples/sql.lp}{sql.lp}}
}
Note that the values from the answer set are inserted into the database.
In fact the output of the program changes when it is consecutively called.
Even more non-determinism can be added by using the option \code{--rand-freq=1.0}
to induce 100 percent random decisions.
\com{Admittedly, this example is somewhat spacy. Better ideas?}
\eexample
\end{example}

>>>>

\index{Statements!Rules}
\index{Statements!Facts}
\index{Statements!Integrity Constraints}
Rules are defined as follows:\\

\begin{tabular}{rl@{}l}\label{eq:normal:rule}
\textbf{Rule:} & $A_0$& $\code{:-}~L_1\code{,}\dots\code{,}L_n$\code{.}
\\
\textbf{Fact:} & $A_0$&\code{.}
\\
\textbf{Integrity Constraint:} & & $\code{:-}~L_1\code{,}\dots\code{,}L_n$\code{.}
\end{tabular}
\\

\noindent
The head~$A_0$ of a fact or a rule is an \emph{atom} of the same form
as a function symbol or constant.
Any $L_j$ is a \emph{literal} of the form $A$ or $\code{not}~A$
for an atom~$A$ where the connective \code{not} corresponds to default negation\index{Literals!Default Negation}.
The set of literals $\{L_1,\dots,L_n\}$ is called the body of the rule.
Facts have an empty body.
Throughout this section we further extend the predicates that can be used 
in a rule including comparison predicates (Section \ref{subsec:gringo:comp}) and aggregates (Section \ref{subsec:gringo:aggregate}).
Furthermore, \gringo{} expects rules to be safe, i.e., \index{Safe Program}
all variables that appear in a rule have to appear in some positive literal (a literal not preceded by \code{not}) in the body.
If a variable appears positively in some predicate, then we say that this predicate binds the variable.

Intuitively, the head of a rule has to be true whenever all its body literals are true.
In ASP every atom needs some derivation, i.e., 
an atom cannot be true if there is no rule deriving it.
This implies that only atoms appearing in some head can appear in answer sets.
Furthermore, derivations%
\footnote{There are extensions like disjunctions 
that go beyond simple derivability and also require minimality w.r.t. a reduct. 
We do not cover the semantics of such constraints in this guide.}%
have to be acyclic,
a feature that is important to model reachability.
As a simple example, consider the program \code{a :- b. b :- a.}
The only answer set to this program is the empty set.
Adding either \code{a.} or \code{b.} to the program results in the answer set $\{a,b\}$.
Finally, note that default negation is ignored when checking for acyclic derivations 
(we do not need a reason for an atom being false).
Default negation can be used to express choices, e.g., the program \code{a :- not b. b :- not a.} 
has the two answer sets $\{a\}$ and $\{b\}$.
But in practice it is never needed to express choices this way.
For example in the introductory example in Section~\ref{sec:quickstart} we used a cardinality 
constraint, which provides a much more readable way to introduce choices.
%\com{This paragraph contains a lot of information. Could someone help me making it more readable?}

@@

In logic programs, connective \code{not} expresses default negation,
that is, a literal $\code{not}~A$ is assumed to hold unless~$A$ is derived.
In contrast, the classical (or strong) negation of some proposition
holds if the complement of the proposition is derived~\cite{gellif91a}.
Classical negation, indicated by symbol ``\code{-},'' is permitted in front of atoms.
That is, if~$A$ is an atom, then~$\code{-}A$ is the complement of~$A$.
Semantically, $\code{-}A$ is simply a new atom,
with the additional condition that~$A$ and~$\code{-}A$ must not jointly hold.
Observe that classical negation is merely a syntactic feature that can be
implemented via integrity constraints whose effect is to eliminate
any answer set candidate containing complementary atoms.

@@

\begin{Note}
that it is important that variables
in the scope of an arithmetic function are not bound by a corresponding atom.
For instance, the rule \code{p(X) :- p(X+1).} is not safe but \code{p(X-1) :- p(X).} is.
Although, the latter might produce an infinite grounding 
and \gringo{} not necessarily halts when given such an input.
\end{Note}

@@

All comparison predicates can also be used with arbitrary ground terms,
as in the next program:%
\marginlabel{%
  As above, invoking:\\
  \code{\mbox{~}gringo -t \attach{examples/symbc.lp}{symbc.lp}}
  yields the unique answer set of the program in terms of facts.}
%
\lstinputlisting{examples/symbc.lp}
%
Integers are compared in the usual way and constants are ordered lexicographically.
Function symbols are compared first using their arity.
If the arity differs, then the name of the function symbol is compared lexicographically.
If again the name differs, then arguments are compared component wise.
Finally, integers are always smaller than constants and constants are always smaller than function symbols.
\eexample
%\comment{add \#sup and \#inf?}
\end{example}

\begin{Note}
that a built-in comparison predicate cannot bind variables, i.e.,
when checking whether a rule is safe, 
comparison predicates are not considered to be positive.
\end{Note}

@@

\begin{Note}
that assignments to some extent can bind variables.
Of course cyclic assignments cannot bind variables.
For example the rule \code{p(X) :- X = Y, Y = X.} is rejected by \gringo{}.
Either \code{X} or \code{Y} has to be provided by some positive predicate in this case.
Additionally, unification is restricted to ground terms on the right hand side of the assignment, 
that is, all variables on the right hand side have to be bound by some other predicate.
\end{Note}

@@

\begin{Note}
that as with built-in arithmetic functions,
an integer interval mentioning some variable (like~\var{X} in Line~4 of Example~\ref{ex:int})
cannot be used to bind the variable.
\end{Note}

@@

\begin{Note}
that there are three important issues about the correct usage of conditions:
\begin{enumerate}
\item All predicates of atoms on the right-hand side of a condition
must be either domain predicates,%\comment{refer to appendix?}
i.e., predicates that can be completely evaluated during grounding,
or built-in, which is due to the fact that conditions are evaluated during grounding.
\item Any variable occurring within a condition is considered as \emph{local},
that is, a condition cannot be used to bind variables outside the condition.
In turn, variables outside conditions are \emph{global}, and each variable
within an atom in front of a condition must occur on the right-hand side or
be global.
\item Global variables take priority over local ones, that is,
they are instantiated first.
As a consequence, a local variable that also occurs globally is substituted by a term
before the ground instances of a condition are determined.
Hence, the names of local variables must be chosen with care,
making sure that they do not accidentally match the names of global variables.
\end{enumerate}
\end{Note}

@@

%\com{Should we mention stratifed aggregates that do not necessarily need conditions?}

@@

\paragraph{Constant Replacement.}
\index{Meta-Statements!Constant Replacement, \code{\#const}}%
Constants appearing in a logic program may actually be placeholders for
concrete values to be provided by a user.
An example of this is given in Section~\ref{subsec:ex:color}.
Via the \code{\#const} declarative,
one may define a default value to be inserted for a constant.
Such a default value can still be overridden via command line option
\code{--const} (cf.\ Section~\ref{subsec:opt:gringo}).
Syntactically, \code{\#const} must be followed by an assignment having
a (symbolic) constant on the left-hand side and a term on the right-hand side.
Some exemplary constant declarations are:
%
\begin{lstlisting}[numbers=none]
#const x = 42.
#const y = f(x,h).
\end{lstlisting}

\begin{Note}
that (for efficiency reasons) constant declarations are order dependent.
In the example above \code{x} would be replaced by \code{42}
but when reversing the directives this would no longer be the case.
\end{Note}

\paragraph{Domain Declarations.}
\index{Meta-Statements!Domain Declarations, \code{\#domain}}%
Usually, variable names are local to a rule,
where they must be bound via appropriate atoms.
This locality can be undermined by using
\code{\#domain} declarations 
that globally associate variable names to atoms.
An associated atom is then simply added to the body of a rule in which
such a predefined variable name occurs in.
The following is a made-up example:
%
\begin{lstlisting}[numbers=none]
p(1,1). p(1,2).
#domain p(X,Y).
#domain p(Y,Z).
q(Z,X) :- not p(Z,X).
\end{lstlisting}
%
The above program is a priori not safe because
variables~\var{X} and~\var{Z} are unbound in the last rule.
However, as they belong to \code{\#domain} declarations,
\gringo\ and \clingo\ expand the last rule to:
%
\begin{lstlisting}[numbers=none]
q(Z,X) :- p(X,Y), p(Y,Z), not p(Z,X).
\end{lstlisting}
%
Observe that the resulting program is safe.

\begin{Note}
that we suggest not to use domain statements because 
in ASP it is common to use very short variable names and
using domain statements likely results in name clashes
and undesired behavior.
\end{Note}

@@

\begin{Note}
that compute statement are not needed in general.
The same behavior can be achieved by using integrity constraints.
In fact, compute statements exist mainly for compatibility reasons with \lparse.
We suggest to not use them.
\end{Note}

@@

\marginlabel{%
Try to invoke this program multiple times using:\\
\code{%
\mbox{~}clingo \attach{examples/sql.lp}{sql.lp}}
}
Note that the values from the answer set are inserted into the database.
In fact the output of the program changes when it is consecutively called.
Even more non-determinism can be added by using the option \code{--rand-freq=1.0}
to induce 100 percent random decisions.
%\com{Admittedly, this example is somewhat spacy. Better ideas?}
\eexample
\end{example}
